Probabilistic Model of Bilateral Lymphatic Spread in Head and Neck Cancer

### 2024-10-18 manuscript.qmd
Built site for gh-pages
Update on Overleaf.

When treating head and neck squamous cell carcinomas (HNSCC) with radiotherapy or surgery, not only the primary tumor and the clinically detected lymph node metastases are targeted. Per current guidelines, large volumes of the neck are part of the the elective clinical target volume (CTV-N) [@gregoire_ct-based_2003;@gregoire_delineation_2014;@gregoire_delineation_2018;@eisbruch_intensity-modulated_2002;@biau_selection_2019;@chao_determination_2002;@vorwerk_guidelines_2011;@ferlito_elective_2009]. This is to minimize the risk of regional recurrences due to untreated microscopic disease which cannot be detected by current in-vivo imaging modalities such as computed tomography (CT), magnetic resonance imaging (MRI), or positron emission tomography (PET). However, minimizing the risk of missing occult disease in the lymph drainage region must be balanced against the toxicity related to unnecessary treatment of healthy tissue.

The aforementioned guidelines for the definition of the CTV-N are based on anatomically defined lymph node levels (LNL) [@gregoire_delineation_2014] and the overall prevalence of lymph node metastases in the LNLs. Based on this, they recommend extensive irradiation of both sides of the neck for a majority of patients. However, this prevalence the guidelines are based on does not take into account an individual patient's state of tumor progression, which may have a great impact on a patient's personal risk for occult disease in a given LNL. For example, a patient with substantial nodal involvement in the ipsilateral side and an advanced primary tumor that crosses the mid-sagittal plane would receive extensive elective irradiation of the contralateral LNLs II, III, and IVa. However, if the patient presented with no clinically detectable nodal disease (cN0) and only a small, clearly lateralized T1 tumor, the same contralateral CTV-N would be recommended [@biau_selection_2019].

To better quantify this personalized risk of occult disease we have previously developed an intuitive, probabilistic hidden Markov model [@ludwig_hidden_2021;@ludwig_modelling_2023] which was originally based on a conceptually similar Bayesian network model [@pouymayou_bayesian_2019]. So far, these models were limited to the ipsilateral side of the neck. The main contribution of this work will be to extend the formalism to include risk predictions for the contralateral side as well. We believe this holds the potential to reduce the electively irradiated volumes in the contralateral neck for patients whose individual risk for occult disease is low. In turn, this may substantially reduce the toxicity from irradiation and the associated decrease in quality of life.

@@

To be able to create models for lymphatic tumor progression that include all relevant LNLs including the contralateral side, we have collected a detailed dataset of {{< var data.num_patients >}} patients with newly diagnosed oropharyngeal squamous cell carcinomas [@ludwig_dataset_2022;@ludwig_multi-centric_2023]. It reports the lymph node involvement per LNL of every patient individually in tabular form, in addition to other primary tumor and patient characteristics such as T-category, subside, and lateralization of the primary tumor, and HPV p16 status. Their patient records have been collected at four different institutions and a brief overview over some of their patients' characteristics are shown in @tbl-data-overview. The data from the Inselspital Bern and the Centre Léon Bérard only consist of patients who received a neck dissection. Since this treatment is more commonly chosen for early T-category patients, they also make up a larger portion of the respective dataset. The majority of patients from the University Hospital Zürich and the Vall d'Hebron Barcelona Hospital was treated with definitive radiotherapy. For a subset of 83 patients in the CLB dataset, the lateralization of the primary tumor was not reported.

@@

Pathological involvement is thus available only for a subset of the patients and only for the dissected levels. For the remaining patients, only clinical involvement is available. For the analysis performed in this paper, all available diagnostic information has been combined into a consensus decision representing the most likely state of involvement for each patient and LNL, using literature values for sensitivity and specificity of different modalities [@de_bondt_detection_2007;@kyzas_18f-fluorodeoxyglucose_2008]. These values are tabulated in @tbl-spec-sens.

This consensus was formed as follows: Suppose the involvement status $X_v$ of LNL $v$ was assessed using different diagnostic modalities $\mathcal{O} = \{ \text{MRI}, \text{CT}, \text{pathology}, \ldots \}$, each characterized by their own pair of sensitivity and specificity values $s_N^{\mathcal{o}}$ and $s_P^{\mathcal{o}}$, with $\mathcal{o} \in \mathcal{O}$. Then we have $|\mathcal{O}|$ observations $z_v^{\mathcal{o}} \in \left[ 0, 1 \right]$, where 0 stands for "healthy" and 1 for "involved". We can then compute the most likely true involvement $X_v$ using the likelihood function

@@

The entire data, including additional patients with tumors in other primary locations than the oropharynx, is publicly available: It may be [downloaded from LyProX](https://lyprox.org/patients/dataset) where it can be interactively explored too, [from GitHub](https://github.com/rmnldwg/lydata), [from zenodo](https://zenodo.org/search?q=lydata), or via the *Data-in-Brief* publications @ludwig_dataset_2022 and @ludwig_multi-centric_2023. The Data-in-Brief publications contain a detailed description of the datasets and the data format. However, these publications do not include the most recent dataset addition from Vall d'Hebron Barcelona Hospital.

@@

These datasets allow us to investigate correlations between the involvement of individual LNLs, or between risk factors and involvement. In @fig-data-strat, we have plotted the prevalence of each contralateral LNL's involvement, stratified by risk factors such as T-category, number of ipsilaterally involved LNLs, and whether the tumor extended over the mid-sagittal line. We also show in the fourth subfigre that T-category and ipsilateral involvement are not just rpedictive of contralateral involvement because they correlate with midline extension. A similar but more complete stratification is also tabulated in the appendix in @tbl-data-strat.

@@

The top left panel in @fig-data-strat indicates that T-category is correlated with contralateral involvement (as it is with overal involvement). This is simply because T-category may on average be considered a surrogate for the time between onset of disease and diagnosis. I.e., a patient with a T4 tumor was -- on average -- diagnosed later than a patient with a T1 tumor. Thus, the former did have more time to develop metastases.

Similarly, ipsilateral involvement correlates with contralateral metastasis. The tumor of a patients with many metastases in ipsilateral LNLs was probably able to spread for longer (or faster) compared to a tumor in a patient with no nodal disease. This, too, may therefore be considered a surrogate for the duration of the disease. In addition, it has been hypothesized that bulky nodal disease ipsilaterally may also redirect lymph fluids to the contralateral side. <!-- citation needed? -->

The bottom left panel in @fig-data-strat shows that patients with a tumor crossing the mid-sagittal line show contralateral involvement vastly more often compared to patients with clearly lateralized tumors. This makes intuitive sense, because the lymphatic system in the head and neck region is typically symmetric and thus no major vessels cross the midline. Therefore, interstitial fluids from the primary tumor -- which we assume to carry living malignant cells -- may only reach the blind-ended lymphatic vessels in the contralateral neck via short-ranged diffusion. Which in turn is only possible when the primary tumor is close enough to the mid-sagittal line or crosses it. <!-- citation(s) needed? or is this intuition good enough? -->

@@

1. More advanced T-category should lead to higher risk for nodal disease. One approach to achieve this via the expected time of diagnosis has already been developed in the form of a hidden Markov model [@ludwig_hidden_2021].
2. The degree of ipsilateral involvement should give the model information on the time that may have passed between onset and diagnosis of the disease. This should come in addition to what can be inferred about the time from T-category alone.
3. A tumor that extends over the mid-sagittal line should yield contralateral metastases with much higher probability.

Over the course of this work, we will first briefly recap the HMM in @sec-unilateral, which was so far used to model ipsilateral lymphatic progression only. Then, we intuitively extend it to include the contralateral side as well in @sec-ext-to-contra. In this section, we also introduce a way of modelling the tumor's midline extension as a random variable (@sec-midline) and lastly talk about how it may affect the contralateral spread in @sec-params-symmetry.

@@

In @tbl-midline-params, we tabulate the mean and standard deviation of the sampled parameters for the full midline model. Considering the parameters of the bilateral model that describe the ipsilateral spread ($b^i_v$, $t_{vr}$ and $p_\text{adv.}$), we note that the bilateral model mostly reproduces the parameter values reported in the earlier publication [@ludwig_modelling_2023] for the unilateral model. The small differences may be mostly due to the differences in training dataset. Therefore, the analysis of the models capability to adequately describe ipsilateral lymph node involvement is not repeated here and we focus instead on the analysis of contralateral involvement. We can further note that the contralateral spread parameters $b^c_v$ are small compared to the ipsilateral parameters $b^i_v$, reflecting the low prevalence of contralateral lymph node involvement for lateralized tumors.

```{python}
#| echo: false
#| label: tbl-midline-params
#| tbl-cap: Mean sampled parameter estimates of the midline model and the respective standard deviation.
def map_param_names(name: str) -> str:
  """Make parameter names more readable."""
  try:
    return {
      "midext_prob": "Mid. ext. probability",
      "mixing": "Mixing ⍺",
      "late_p": "late T-cat. binom. prob.",
    }[name]
  except KeyError:
    return re.sub(
      r"(ipsi|contra)_", r"\g<1>: ", name
    ).replace(
      "to", " ➜ "
    ).replace(
      "_spread", ""
    )

@@

The model introduced in the previous section defines a model of the joint probability distribution over midline extension and ipsi- and contralateral lymph node involvement, $P \left( \mathbf{X}^\text{i}, \mathbf{X}^\text{c}, \epsilon \right)$. How this quantity is computed is illustrated in @fig-model-state-dist, visually representing @eq-bilateral-marginal-algebra. To keep it manageable to interpret, this example only considers the LNLs II, III, and IV ipsi- and contralaterally, resulting in $2^3 = 16$ distinct states per side and thus $2 \times 8 \times 8 = 128$ total states. LNLs I, V, and VII and the associated spread parameters have been removed from the model while the remaining parameters are set to the mean values in table xxx. @fig-model-state-dist shows the joint distribution as two separate heatmaps for the two midline extension states for early/advanced??? T-category tumors. The most likely state is a lateralized tumor with involvement of ipsilateral level II and no contralateral involvement, which has a probability of approximately 25%. The second most probable state is a lateralized tumor with ipsilateral involvement of levels II and III and no contralateral involvement. The most likely state with contralateral involvement are tumors with midline extension with involvement of contralateral level II and ipsilateral level II and III.

![Visual representation of @eq-bilateral-marginal-algebra for the case of midline extention. The left and right matrix represent the evolution of the possible hidden states for the ipsi- and contralateral neck respectively. In particular, the right matrix describes the evolution of the contralateral hidden states *and* midline extension. In the center, the time-prior is plotted as a diagonal matrix. The result of this matrix computation is the joint distribution $P \left( \mathbf{X}^\text{i}, \mathbf{X}^\text{c}, \epsilon=\texttt{True} \right)$](figures/fig_model_state_dist.svg){#fig-model-state-dist}

@@

![Comparison of predicted (histograms) vs observed (beta posteriors) prevalences. Shown for the contralateral LNLs II (blue), III (orange), and IV (green). The top row shows scenarios with early T-category tumors, the bottom row for late T-category ones. The left column depicts scenarios where the primary tumor is clearly lateralized, the right column scenarios of tumors extending over the mid-sagittal line. This figure illustrates the model's ability to describe the prevalence of involvement for different combinations of the risk factors T-category and midline extension.](figures/fig_model_prevalences_overall.svg){#fig-model-prevalences-overall}

@Fig-model-prevalences-overall shows nicely that the model is capable of accurately accounting for the most important risk factors, i.e. T-category and midline extension. As observed in the data, the model predicts that the prevalence of contralateral LNL II involvement increases from {{< var prevalence.early_noext_iN0_cII.predicted >}}% for early T-category lateralized tumors to {{< var prevalence.late_ext_iN0_cII.predicted >}}% when the tumor is of advanced T-category and crosses the mid-sagittal line. Similarly, the prevalence of contralateral LNL III involvement increases from around {{< var prevalence.early_noext_iN0_cIII.predicted >}}% for early T-category lateralized tumors to almost {{< var prevalence.late_ext_iN0_cIII.predicted >}}% when the tumor is of advanced T-category and crosses the mid-sagittal line.

@@

![Comparison of the computed and observed prevalences for scenarios that illustrate the model's capability of accounting for the correlation between ipsi- and contralateral involvement. We show two scenarios: One where the ipsilateral neck shows no involvement (at least LNLs I to V are healthy, LNL VII was ignored because data on it is missing for some patients) in blue and one where at least LNL III was involved in orange. These two scenarios are plotted for all combinations of T-category (early in top row, advanced in bottom row) and tumor lateralization (lateralized in left column, extending over mid-sagittal line in right column).](figures/fig_model_prevalences_with_ipsi.svg){#fig-model-prevalences-with-ipsi}

In @fig-model-prevalences-with-ipsi we display the model's ability to capture the correlation between ipsi- and contralateral involvement. It shows that the prevalence of metastases in the two sides of the neck is correlated via the time of diagnosis, despite the model not having any direct connections between the two side. However, there are some discrepancies in the model's prediction: It cannot quite capture the correlation of ipsi- and contralateral involvement that is seen in the data. This is notable for example in the bottom right subplot where the prevalence is overestimated for the scenario of an (almost) healthy ipsilateral neck and underestimated when at least LNL III shows metastases. One reason for this may be the width of the time-prior: If it was wider, the posterior over the diagnosis time given the ipsilateral involvement could more flexibly shift to reflect a more severe contralateral involvement.

@@

@@

![Histograms over the predicted risk of occult contralateral LNL II (left), III (middle), and IV (right) invovlement, shown for some combinations of T-category, tumor lateralization, and clinical LNL diagnoses. All LNLs not explicitly mentioned in the legend, including the LNL for which the risk of occult disease was computed, where assumed to be clinically negative, based on CT imaging (specificity 76%, sensitivity 81%).](figures/fig_model_risks.svg){#fig-model-risks}

@@

@Fig-model-risks (left panel) shows the predicted risk of occult disease in contralateral LNL II. The most important variable impacting the prediction for contralateral level II involvement in our model is the tumor's lateralization. A patient with a clearly lateralized early T-category tumor and a clinically N0 neck is predicted to have a 1-2% <!-- TODO: make this dynamic --> risk for occult disease in contralateral LNL II. For a patient with a clearly lateralized early T-category tumor but *with* mid-sagittal extension, the risk increases to {{< var risk.early_ext_iN0_cN0_II >}}%.

Also advanced T-category increases the risk of occult disease but plays a lesser role: Considering the scenario of an early T-category tumor that crosses the midline and an ipsilateral neck where the LNLs II and III are clinically involved, the risk for occult contralateral LNL II disease is around {{< var risk.early_ext_iII+III_cN0_II >}}%. For the same scenario and an advanced T-category tumor, the risk increases to {{< var risk.late_ext_iII+III_cN0_II >}}%.

Lastly, the predicted risk is also correlated via the time-steps to the degree of ipsilateral involvement. When an early T-category patient with a tumor that extends over the mid-sagittal line presents with a clinically N0 ipsilateral neck instead of involvement in the levels II and III, the risk is {{< var risk.early_ext_iN0_cN0_II >}}% instead of {{< var risk.early_ext_iII+III_cN0_II >}}%.

Taken together, T-category and ipsilateral involvement may still considerably impact the risk prediction for contralateral involvement: In @fig-model-risks the scenarios underlying the orange ({{< var risk.early_ext_iN0_cN0_II >}}%) and the purple ({{< var risk.late_ext_iII+III_cN0_II >}}%) histograms differ in T-category (early vs advanced) and clinical diagnosis (N0 vs ipsilateral LNL II and III involved).

@@

As shown in the center subplot of @fig-model-risks, the risk for occult disease in the contralateral LNL III may only cross a 5% threshold if the upstream level II is clinically involved, too. If the clinical diagnosis of the contralateral upstream level II is pathologically confirmed by FNA the risk increases by another percentage point to {{< var risk.late_noext_iII_cII_FNA_III >}}%. When the tumor additionally crosses the mid-sagittal line, the risk may reach around {{< var risk.late_ext_iII_cII_FNA_III >}}%.

@@

Our model's prediction for the contralateral LNL IV show that even in the most extreme case, with an advanced T-category tumor extending over the mid-sagittal line as well as metastases in the ipsilateral LNLs II-IV and the contralateral upstream levels II and III, the risk stays at just {{< var risk.late_ext_iII+III+IV_cII+III_IV >}}%. We only observe a risk of {{< var risk.late_ext_iII+III+IV_cII+III_FNA_IV >}}% when the involvement of the upstream level III is pathologically confirmed after taking a biopsy using FNA. This is because the model's prior probability for contralateral LNL IV involvement is so low that even given a clinical diagnosis in that level, a false positive in LNL III is a more likely explanation. This possibility, however, drops virtually to zero when tumor cells are pathologically confirmed in level III.

@@

Even for advanced T-category patients whose tumor crosses the midline and who present with extensive ipsi- and contralateral clinical involvement, the risk for involvement in the LNLs I and VII never exceeds 3%. LNL V contralaterally also shows only a very low risk for occult disease, but it does rise to {{< var risk.late_ext_iI+II+III+IV+V+VII_cII+III+IV_FNA_V >}}% in the same extreme case when additionally the upstream level IV involvement has been confirmed with FNA.

@@

The predictions of this model are being used to inform a clinical trial on volume deescalation at the University Hospital Zurich [citation needed<!-- TODO: add -->]. When accepting a 5% risk of occult disease in any given LNL, the model suggests the following contralateral elective irradiation (assuming the respective LNL appears clinically healthy):

>>>>

When treating head and neck squamous cell carcinomas (HNSCC) with radiotherapy or surgery, not only the primary tumor and the clinically detected lymph node metastases are targeted. Per current guidelines, large volumes of the neck are part of the elective clinical target volume (CTV-N) [@gregoire_ct-based_2003;@gregoire_delineation_2014;@gregoire_delineation_2018;@eisbruch_intensity-modulated_2002;@biau_selection_2019;@chao_determination_2002;@vorwerk_guidelines_2011;@ferlito_elective_2009]. This is to minimize the risk of regional recurrences due to untreated microscopic disease which cannot be detected by current in-vivo imaging modalities such as computed tomography (CT), magnetic resonance imaging (MRI), or positron emission tomography (PET). However, minimizing the risk of missing occult disease in the lymph drainage region must be balanced against the toxicity related to unnecessary treatment of healthy tissue.

The aforementioned guidelines for the definition of the CTV-N are based on anatomically defined lymph node levels (LNL) [@gregoire_delineation_2014] and the overall prevalence of lymph node metastases in the LNLs. Based on this, they recommend extensive irradiation of both sides of the neck for a majority of patients. However, the overall prevalence of involvement in a given LNL does not equal a patient's personal risk for occult disease in that LNL, which depends on an individual patient's state of tumor progression. Currently, a patient presenting with no clinically detectable nodal disease (cN0) and a small, clearly lateralized T1 tumor, the same recommended contralateral CTV-N as a patient with substantial nodal involvement in the ipsilateral side and an advanced primary tumor that crosses the mid-sagittal plane. Both patients receive extensive elective irradiation of the contralateral LNLs II, III, and IVa [@biau_selection_2019].

To better quantify this personalized risk of occult disease we have previously developed an intuitive, probabilistic hidden Markov model [@ludwig_hidden_2021;@ludwig_modelling_2023] which was originally based on a conceptually similar Bayesian network model [@pouymayou_bayesian_2019]. So far, these models were limited to the ipsilateral side of the neck. The main contribution of this work is the extension of the formalism to include risk predictions for the contralateral side as well. The model may then guide the reduction of the electively irradiated volumes in the contralateral neck for patients whose individual risk for occult disease is low. In turn, this may substantially reduce the toxicity from irradiation and the associated decrease in quality of life.

@@

To be able to create models for lymphatic tumor progression that include all relevant LNLs including the contralateral side, we have collected a detailed dataset of {{< var data.num_patients >}} patients with newly diagnosed oropharyngeal squamous cell carcinomas [@ludwig_dataset_2022;@ludwig_multi-centric_2023]. It reports the lymph node involvement per LNL of every patient individually in tabular form, in addition to other primary tumor and patient characteristics such as T-category, subsite, lateralization of the primary tumor, and HPV p16 status. Their patient records have been collected at four different institutions and a brief overview over some of their patients' characteristics are shown in @tbl-data-overview. The data from the Inselspital Bern (ISB) and the Centre Léon Bérard (CLB) only consist of patients who all received a neck dissection. The majority of patients from the University Hospital Zürich (USZ) and the Vall d'Hebron Barcelona Hospital (VBH) was treated with definitive radiotherapy. Since sugical treatment is more common for early T-category patients, the ISB and CLB datasets contain a larger portion of early T-category patients compared to the USZ and VBH dataset. For a subset of 83 patients in the CLB dataset, the lateralization of the primary tumor was not reported.

@@

Pathological involvement is available only for the subset of surgical patients and only for the dissected levels. For the remaining patients, only clinical involvement is available. For the analysis performed in this paper, all available diagnostic information has been combined into a consensus decision representing the most likely state of involvement for each patient and LNL, using literature values for sensitivity and specificity of different modalities [@de_bondt_detection_2007;@kyzas_18f-fluorodeoxyglucose_2008]. How this consensus is formed is described in detail in appendix XXX. In brief, pathology after neck dissection is taken as the ground truth if available, overwriting any contradicting clinical involvement diagnoses. If a level was not dissected, the most likely state is determined by PET-CT in most cases.

@@

The entire data, including additional patients with tumors in other primary locations than the oropharynx, is publicly available: It may be [downloaded from LyProX](https://lyprox.org/patients/dataset) where it can be interactively explored too, [from GitHub](https://github.com/rmnldwg/lydata), [from zenodo](https://zenodo.org/search?q=lydata), or via the *Data-in-Brief* publications @ludwig_dataset_2022 and @ludwig_multi-centric_2023. The Data-in-Brief publications contain a detailed description of the datasets and the data format. These publications do not include the most recent dataset addition from Vall d'Hebron Barcelona Hospital, which will be described in a future publication.

@@

These datasets allow us to investigate correlations between the involvement of individual LNLs, or between risk factors and involvement. In @fig-data-strat, we have plotted the prevalence of each contralateral LNL's involvement, stratified by the risk factors T-category, number of ipsilaterally involved LNLs, and whether the tumor extended over the mid-sagittal plane.

@@

The bottom left panel in @fig-data-strat shows that patients with a tumor crossing the mid-sagittal plane show contralateral involvement vastly more often compared to patients with clearly lateralized tumors. This makes intuitive sense, because the lymphatic system in the head and neck region is typically symmetric and no major lymph vessels cross the midline. Therefore, interstitial fluids from the primary tumor -- which we assume to carry living malignant cells -- may only reach the blind-ended lymphatic vessels in the contralateral neck via short-ranged diffusion. Which in turn is only possible when the primary tumor is close enough to the mid-sagittal line or crosses it. <!-- citation(s) needed? or is this intuition good enough? -->

The top left panel in @fig-data-strat indicates that T-category is correlated with contralateral involvement (as it is with overal involvement). This is because T-category may be considered a surrogate for the time between onset of disease and diagnosis. I.e., a patient with a T4 tumor was -- on average -- diagnosed later than a patient with a T1 tumor. Thus, the former did have more time to develop metastases.

Similarly, ipsilateral involvement correlates with contralateral metastasis (top right panel). The tumor of a patient with many metastases in ipsilateral LNLs was probably able to spread for longer (or faster) compared to a tumor in a patient with no nodal disease. The degree of ipsilateral involvement, too, may therefore be considered a surrogate for the duration of the disease. In addition, it has been hypothesized that bulky nodal disease ipsilaterally may also redirect lymph fluids to the contralateral side. <!-- citation needed? -->

@@

1. A tumor that extends over the mid-sagittal line should yield contralateral metastases with much higher probability.
2. More advanced T-category should lead to higher risk for nodal disease. One approach to achieve this via the expected time of diagnosis has already been developed in the form of a hidden Markov model [@ludwig_hidden_2021].
3. The degree of ipsilateral involvement should give the model information on the time that may have passed between onset and diagnosis of the disease. This should come in addition to what can be inferred about the time from T-category alone.

<!-- Over the course of this work, we will first briefly recap the HMM in @sec-unilateral, which was so far used to model ipsilateral lymphatic progression only. Then, we intuitively extend it to include the contralateral side as well in @sec-ext-to-contra. In this section, we also introduce a way of modelling the tumor's midline extension as a random variable (@sec-midline) and lastly talk about how it may affect the contralateral spread in @sec-params-symmetry. -->

@@

In @tbl-midline-params, we tabulate the mean and standard deviation of the sampled parameters for the full midline model. Considering the parameters of the bilateral model that describe the ipsilateral spread ($b^i_v$, $t_{vr}$ and $p_\text{adv.}$), we note that the bilateral model mostly reproduces the parameter values reported in the earlier publication [@ludwig_modelling_2023] for the unilateral model. The small differences may be mostly due to the differences in the training dataset. Therefore, the analysis of the model's capability to adequately describe ipsilateral lymph node involvement is not repeated here and we focus instead on the analysis of contralateral involvement. We can further note that the contralateral spread parameters $b^c_v$ are small compared to the ipsilateral parameters $b^i_v$, reflecting the low prevalence of contralateral lymph node involvement for lateralized tumors.

```{python}
#| echo: false
#| label: tbl-midline-params
#| tbl-cap: Mean sampled parameter estimates of the midline model and the respective standard deviation. The parameters set to fixed values are the maximum number of time steps $t_{max}=10$ and the time prior parameter for early T-category patients $p_{early}=0.3$.
def map_param_names(name: str) -> str:
  """Make parameter names more readable."""
  try:
    return {
      "midext_prob": "Mid. ext. probability",
      "mixing": "Mixing ⍺",
      "late_p": "late T-cat. binom. prob.",
    }[name]
  except KeyError:
    return re.sub(
      r"(ipsi|contra)_", r"\g<1>: ", name
    ).replace(
      "to", " ➜ "
    ).replace(
      "_spread", ""
    )

@@

The model introduced in the previous section defines a model of the joint probability distribution over midline extension and ipsi- and contralateral lymph node involvement, $P \left( \mathbf{X}^\text{i}, \mathbf{X}^\text{c}, \epsilon \right)$. How this quantity is computed is illustrated in @fig-model-state-dist, visually representing @eq-bilateral-marginal-algebra. To keep it manageable to interpret, this example only considers the LNLs II, III, and IV ipsi- and contralaterally, resulting in $2^3 = 16$ distinct states per side and thus $2 \times 8 \times 8 = 128$ total states. LNLs I, V, and VII and the associated spread parameters have been removed from the model while the remaining parameters are set to the mean values in table XXX.

The matrix on the left side of @fig-model-state-dist shows the time evolution of the probability distribution over the ipsilateral involvement state, starting from the healthy state $[0,0,0]$ with probability one. The two matrices on the right side shows the time evolution of the probability distribution over the contralateral involvement state, where we need to distinguish the two cases with and without midline extension. At time step $t=0$, the contralateral neck starts from the healthy state $[0,0,0]$ without midline extension. The matrix in the center shows the time prior for EARLYorLATE T-category. The result of the matrix multiplication is the joint probability distribution over midline extension and ipsi- and contralateral lymph node involvement, $P \left( \mathbf{X}^\text{i}, \mathbf{X}^\text{c}, \epsilon=\texttt{True} \right)$ shown in figure XXX. The joint distribution is shown as two separate heatmaps for the two midline extension states. The most likely state is a lateralized tumor with involvement of ipsilateral level II and no contralateral involvement, which has a probability of approximately 25%. The second most probable state is a lateralized tumor with ipsilateral involvement of levels II and III and no contralateral involvement. The most likely state with contralateral involvement are tumors with midline extension with involvement of contralateral level II and ipsilateral level II and III.

![Visual representation of @eq-bilateral-marginal-algebra for the case of midline extention. The left and right matrix represent the evolution of the possible hidden states for the ipsi- and contralateral neck respectively. In particular, the right matrix describes the evolution of the contralateral hidden states *and* midline extension. In the center, the time-prior for EARLYorLATE T-category is plotted as a diagonal matrix. The result of this matrix computation is the joint distribution $P \left( \mathbf{X}^\text{i}, \mathbf{X}^\text{c}, \epsilon=\texttt{True} \right)$](figures/fig_model_state_dist.svg){#fig-model-state-dist}

@@

![Comparison of predicted (histograms) vs observed (beta posteriors) prevalences, shown for the contralateral LNLs II (blue), III (orange), and IV (green). The top row shows scenarios with early T-category tumors, the bottom row for late T-category ones. The left column depicts scenarios where the primary tumor is clearly lateralized, the right column scenarios of tumors extending over the mid-sagittal line. This figure illustrates the model's ability to describe the prevalence of involvement for different combinations of the risk factors T-category and midline extension.](figures/fig_model_prevalences_overall.svg){#fig-model-prevalences-overall}

@Fig-model-prevalences-overall shows that the model is capable of accurately accounting for the most important risk factors, i.e. T-category and midline extension. As observed in the data, the model predicts that the prevalence of contralateral LNL II involvement increases from {{< var prevalence.early_noext_iN0_cII.predicted >}}% for early T-category lateralized tumors to {{< var prevalence.late_ext_iN0_cII.predicted >}}% when the tumor is of advanced T-category and crosses the mid-sagittal line. Similarly, the prevalence of contralateral LNL III involvement increases from around {{< var prevalence.early_noext_iN0_cIII.predicted >}}% for early T-category lateralized tumors to almost {{< var prevalence.late_ext_iN0_cIII.predicted >}}% when the tumor is of advanced T-category and crosses the mid-sagittal line.

@@

@@

![Comparison of the computed and observed prevalences for scenarios that illustrate the model's capability of accounting for the correlation between ipsi- and contralateral involvement. We show three scenarios where we consider the joint involvement of contralateral LNL II together with different ipsilateral involvements: 1) the ipsilateral neck shows no involvement in blue (LNLs I to V are healthy, LNL VII is unspecified because data on it is missing for some patients), 2) where ipsilateral LNL II is involved in orange (LNLs I, III, IV, and V are healthy), and 3) where ipsilateral LNLs II and III are involved in orange (LNLs I, IV, and V are healthy). These two scenarios are plotted for all combinations of T-category (early in top row, advanced in bottom row) and tumor lateralization (lateralized in left column, extending over mid-sagittal line in the right column).](figures/fig_model_prevalences_with_ipsi.svg){#fig-model-prevalences-with-ipsi}

In @fig-model-prevalences-with-ipsi we display the model's ability to capture the correlation between ipsi- and contralateral involvement. Shown are marginals of the joint distribution where we consider contralateral LNL II involvement with simulaneous involvement of different ipsilateral LNLs. The model has no direct connections between the two sides and thus no adjustable parameter to specifically quantify correlations between ipsi- and contralateral involvement. Metastases in the two sides of the neck are only correlated via the time of diagnosis. Nevertheless, the model adequately predicts that contralateral level II involvement is rare in combination with a healthy ipsilateral neck (green histograms). If ipsilateral LNL II is involved, contralateral involvement is more frequent.

<!-- However, there are some discrepancies in the model's prediction: It cannot quite capture the correlation of ipsi- and contralateral involvement that is seen in the data. This is notable for example in the bottom right subplot where the prevalence is overestimated for the scenario of an (almost) healthy ipsilateral neck and underestimated when at least LNL III shows metastases. One reason for this may be the width of the time-prior: If it was wider, the posterior over the diagnosis time given the ipsilateral involvement could more flexibly shift to reflect a more severe contralateral involvement. -->

@@

![Histograms over the predicted risk of occult involvement in contralateral LNL II (left), III (middle), and IV (right), shown for various combinations of T-category, tumor lateralization, and clinical LNL diagnoses. All LNLs not explicitly mentioned in the legend, including the LNL for which the risk of occult disease was computed, where assumed to be clinically negative (specificity 76%, sensitivity 81%).](figures/fig_model_risks.svg){#fig-model-risks}

@@

@Fig-model-risks (left panel) shows the predicted risk of occult disease in contralateral LNL II. The most important variable impacting the prediction for contralateral level II involvement in our model is the tumor's lateralization. A patient with a clearly lateralized early T-category tumor and clinically detected metastases in ipsilateral LNL II is predicted to have a 1-2% <!-- TODO: make this dynamic --> risk for occult disease in contralateral LNL II (green histogram). For a cN0 patient with a clearly lateralized early T-category tumor but *with* mid-sagittal extension, the risk increases to {{< var risk.early_ext_iN0_cN0_II >}}% (orange histogram).

Also advanced T-category increases the risk of occult disease but plays a lesser role. Considering the scenario of an early T-category tumor that crosses the midline and an ipsilateral neck where the LNLs II and III are clinically involved, the risk for occult contralateral LNL II disease is around {{< var risk.early_ext_iII+III_cN0_II >}}% (red histogram). For the same scenario and an advanced T-category tumor, the risk increases to {{< var risk.late_ext_iII+III_cN0_II >}}% (purple histogram).

Lastly, the predicted risk of occult contralateral LNL II involvement also increases with the degree of ipsilateral involvement. When an early T-category patient with a tumor that extends over the mid-sagittal line presents with a clinically N0 ipsilateral neck instead of involvement in the levels II and III, the risk is {{< var risk.early_ext_iN0_cN0_II >}}% instead of {{< var risk.early_ext_iII+III_cN0_II >}}%.

Taken together, for lateralized tumors, the risk of occult contralateral involvement remains low even for patients with advanced T-category tumors and ipsilateral involvement of LNLs II and III (blue histogram), making midline extension the main risk factor. However, advanced T-category and severe ipsilateral involvement may still considerably increase the predicted risk for contralateral involvement.

@@

As shown in the center subplot of @fig-model-risks, the risk for occult disease in the contralateral LNL III may only cross a 5% threshold if the upstream level II is clinically involved. Even for advanced T-category tumors extending over the midline with extensive ipsilateral involvement (green histogram), the risk of occult metastases in contralateral LNL III is only 2% <!-- TODO: make this dynamic --> if the upstream LNL II is clinically negative. For an advanced T-category tumors extending over the midline with clinical involvement of both ispilateral and contralateral LNL II involvement, the predicted risk in contralateral LNL III increases to 4-5% <!-- TODO: make this dynamic --> (blue histogram). If the clinical diagnosis of the contralateral upstream level II is pathologically confirmed by FNA the risk increases to {{< var risk.late_ext_iII_cII_FNA_III >}}% (red histogram). This is because the possibility that the clinical involvement in contralateral LNL II is a false positive diagnoses is almost excluded. If contralateral LNL II is involved, the predicted risk in contralateral LNL III is {{< var risk.late_noext_iII_cII_FNA_III >}}% even for lateralized tumors (orange histogram).

@@

Our model's prediction for the risk in contralateral LNL IV is below 1% <!-- TODO: make this dynamic --> if the upstream LNL III is clinically negative, that even in the most extreme case of an advanced T-category tumor extending over the mid-sagittal line with metastases in the ipsilateral LNLs II-IV and the contralateral level II (green histogram). If additionally contralateral LNL III is clinically involved, the risk increases to {{< var risk.late_ext_iII+III+IV_cII+III_IV >}}% (blue histogram). Only when the involvement of the upstream level III is pathologically confirmed using FNA, we predict a substantially higher risk of {{< var risk.late_ext_iII+III+IV_cII+III_FNA_IV >}}% (orange histogram). This is because the prior probability for contralateral LNL III and IV involvement is so low that, even given a clinically positive diagnosis in LNL III, it is likely that the involvement in LNL III is a false possitive diagnosis both LNLs III and IV are negative. This possibility, however, is virtually excluded when the involvement in contralateral level III is confirmed by FNA.

@@

Even for advanced T-category patients whose tumor crosses the midline and who present with extensive ipsi- and contralateral clinical involvement, the risk for involvement in the LNLs I and VII never exceeds 3%. Contralateral LNL V also shows only a very low risk for occult disease. Only in the extreme case of severe contralateral involvement including LNL IV, confirmed by FNA, the predicted risk is {{< var risk.late_ext_iI+II+III+IV+V+VII_cII+III+IV_FNA_V >}}%.

@@

The predictions of this model are being used to inform a clinical trial on volume deescalation at the University Hospital Zurich [clinicaltrials.gov, NCT06563362 (DeEscO)<!-- TODO: add -->]. When accepting a 5% risk of occult disease in any given LNL, the model suggests the following contralateral elective irradiation (assuming the respective LNL is clinically healthy):
